---
title: "Specimen Records Reveal Recent Elevational Range Shifts in Puerto Rican *Anolis*"
author: "CJ Battey, Ray Huey, (?)"
date: "May 20, 2016"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(tidy = T)
library(data.table);library(raster);library(ggplot2);library(plyr);library(foreach);library(stringr);
library(magrittr);library(dismo);library(rgeos)
ext.pr <- extent(-67.6,-65.2,17.8,18.7)
ext.yunque <- extent(825000,850000,2015000,2040000)
proj4.wgs <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
proj4.utm <- "+proj=utm +zone=19 +ellps=WGS84 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"

#load elevation rasters
alt <- raster("~/Dropbox/anolis/data/alt_30s_bil/alt.bil") %>% crop(ext.pr) %>% 
  projectRaster(.,projectExtent(.,proj4.utm))
alt1s<- raster("~/Dropbox/anolis/data/elevation/PuertoRico_altitude_1sec_USGSNED2013.tif") %>% crop(ext.pr) %>% 
  projectRaster(.,projectExtent(.,proj4.utm))
```
#Introduction (draft) 
Tropical montane species are thought to be particularly vulnerable to extinction under climate change because they occupy relatively narrow thermal niches and have a finite amount of higher-elevation habitat to occupy in the future. Simple models relating species' ranges to climatic conditions assume that as air temperatures increase, montane species will move uphill to track a thermal niche. However, physiological studies have found that body temperatures in many ectotherms are regulated via fine-scale habitat selection and microclimatic variation between (for example) forested and open habitats. Changes in average temperatures over time or along an altitudinal transect may thus be secondary to local variation in microclimate caused by shifts in land-use or habitat structure when predicting shifts in species' distributions. 

Reforestation of former agricultural lands in Puerto Rico over the course of the 20th century offers an opportunity to compare the impacts of changes in air temperatures and land use in driving shifts in species' ranges. Lizards in the genus *Anolis* are a model system in the fields of ecomorphology, island biogeography, and thermal physiology, and have been studied intensively in Puerto Rico since the early 20th century. *A. gundlachi* is a forest specialist historically considered a montane species, while *A. cristatellus* is a habitat generalist found primarily in open areas at low elevations. Anecdotal reports suggest that *gundlachi* is now thriving at low-elevation sites that were inhabited only by *cristatellus* during the early 1970's, but no published study has assessed shifts in *Anolis* elevational range quantitatively. 

We used aggregated specimen records from natural history museums to assess changes in the distributions of these species in Puerto Rico during the twentieth century. Specifically, we ask: (1) Has *A. gundlachi* expanded to lower elevations since the 1970's? and (2) are range shifts in *A. gundlachi* (if any) better explained by changes in climate or landcover?  

#Methods
#*Anolis* Occurrence Data
####Species Occurrence Points
We downloaded all records of natural history specimens in the genus *Anolis* collected in Puerto Rico from the Global Biodiversity Information Facility (GBIF) and developed a set of scripts to (1) parse ambiguous date formats (2) add verbatim elevations when these were included in the text locality and (3) identify specimens with field-collected GPS coordinates. We then manually georeferenced all collection localities with greater than ten specimen collections using the Google Maps API to measure road kilometers and Google Earth to locate point localities. We followed the vertNet.org best-practices for georeferencing (CITE) and calculated uncertainty to the nearest 100M for all points. All localities were georeferenced by the same individual using the same protocol (supplement 1). 
  
After removing localities with greater than 2km georeferencing uncertainty and splitting out sites with verbatim elevations provided by the collector, we estimated the elevation of each collection as the mean elevation across the full uncertainty radius in the USGS National Elevation Database (2013) at 1-arc-second resolution, using the rgeos and Raster packages in R (CITE x3). Specimen records were binned into four age classes covering the period 1935-2015, based on the forest age ranges given in Helmer et al. (2008)'s analysis of landcover change on Puerto Rico (see below). We subset records to include only species with greater than 200 specimen records before and after 1980 for downstream analyses. 

[ Georeferencing supplement: Although roughly two-thirds of the GBIF records included georeferenced coordinates submitted by natural history museums, we found several clear errors in these data and noted that many cited paper maps, resulting in relatively poor resolution (particularly in the case of road kilometer measurements). To locate each specimen at the finest level possible given existing data, we extracted unique text locality descriptions for all collecting localities (n=1,115) and manually georeferenced sites lacking field-collected GPS coordinates and with greater than five specimen records using Google Earth and Google Maps, following the protocol described in the vertNet.org specimen georeferencing guide. 
 
Briefly, point localities were located through text searches for named places, with an uncertainty radius covering the approximate area of the named place. Coordinates of localities given in road kilometers were assigned by using the Google Maps directions tool to measure road distances from the relevant intersection, with uncertainty corresponding to the significant figures in the text. For localities given in a direction-and-distance format without a road or place name, we calculated road miles when a single clear route existed (e.g. PR 191 south of Palmer) and air miles when multiple road routes were possible, or when the same collector had previously explicitly noted a locality in air miles. Uncertainty for air mile localities was conservatively set to half the distance from the given place name (i.e. start of measurement). ]
 
####Shifts in Elevation Distribution
We found the absolute elevation ranges of each species in each age class, then used a Wilcoxon rank sum test (CITE) to ask whether the elevation distribution of specimen collections had shifted between the periods 1952-1977 and 1991-2015. To test for shifts in the distribution of overall *Anolis* collecting effort between time periods, we then applied a Wilcoxon rank sum test to the full six-species dataset. These analyses were conducted on both a per-specimen and per-locality basis. In the per-locality analysis we subset the data to include only localites with greater than 10 specimens, coded localities as presence/absence for each species, and weighted all localities equally in the analysis. In contrast the per-specimen analysis treats each specimen as an independent data point, which potentially allows greater resolution of shifts in abundance within a preexisting range, but may be biased by heavy sampling of targeted species for specific projects. 

To assess changes in species composition at low elevations specifically, we extracted specimen reports with elevations lower than 250M, split by species, and used a McNemar test (CITE) to compare the relative abundance of each species across time periods (i.e. the number of specimens of species *i* relative to the number of all *Anolis* specimens in each period). This procedure is conceptually similar to the specimen-derived abundance index proposed in Rohwer et al. (Cite YR, Linck et al. 2016, Grayson et al. YR), in which the observed abundance of a target species is corrected for survey effort by dividing by the number of specimens collected with similar techniques in a given area. 
   
Finally, we identified areas which had been surveyed in multiple time periods by calculating pairwise geographic distances among all collecting localities and using a complete-linkage heirarchical clustering algorithm (hclust(method="complete") in R) to identify groups of localities within approximately 1 kilometer of each other. We then merged reports in these locality groups and retained only groups with at least 100 specimen collections each in the periods 1955-1977 and 1991-2015. We then calculated the relative abundance of each species in each locality group and time period, and used a McNemar test to ask if the proportions of each species changed significantly between time periods. 
 
####Landcover
To assess changes in forest cover, we used a forest age raster layer developed via analysis of historic aerial and satellite imagery (Helmer et al. 2008). We merged forest ages across soil types and subset the original Helmer raster layer to produce maps of forested areas of Puerto Rico at 30M resolution in four time bins: 1935-1951, 1952-1977, 1977-1990, and 1991-2000. All areas were considered to be nonforested prior to the maximum age of forests in the Helmer dataset. All analyses GIS and statistical analyses were conducted in R. An alternate version of this manuscript formatted as an RMarkDown document with embedded code including all data processing and analysis is available in the online supplement. 

#Results
####Specimen Database
The final dataset includes 9,680 specimens across 6 species: *A. cristatellus*, *A. gundlachi*, *A. krugi*, *A. evermanni*, and *A. pulchellus*. Most collecting occurred during the periods 1952-1977 or 1991-2016 (Figure 1). The largest institutional collections included in the final dataset are from the Museum of Comparative Zoology at Harvard University and the Museum of Vertebrate Zoology at UC Berkeley. For specimens with georeferenced coordinates, the median distance between GBIF coordinates and our georeferenced location was 0.5 kilometers and the mean was 3.5 kilometers; suggesting that most georeferencing error occured in a relatively small proportion of specimens. 

```{r message=F,fig.height=4,warning=F}
setwd("/R/anolis/")
anolis <- read.csv("~/Dropbox/anolis/data/anolis_data.csv") %>% subset(!is.na(alt))
```

```{r fig.height=2.5}
#distribution of collections across museums
museums <- ddply(anolis,.(institutionCode,timebin),summarize,Reports=length(species))
ggplot(data=museums,aes(x=substr(institutionCode,1,4),y=Reports,fill=timebin))+theme(axis.text.x = element_text(angle = 90, hjust =1,vjust=.5,size=6))+xlab("")+
  geom_bar(stat="identity")+ylab("Specimens")+ggtitle("Specimens  per Institution")
```

```{r fig.height=2.5}
years <- ddply(subset(anolis,year>1960),.(year,species),summarize,Reports=length(alt))
ggplot(data=years,aes(x=year,y=Reports,fill=species))+theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  geom_bar(stat="identity")+ylab("Specimens")+xlab("")
```

Fig 2-3. Specimen source and time summaries.  

####Elevation Ranges
Analyzed on a per-specimen basis, the elevation distributions of all six common *Anolis* species shifted significantly between the periods 1955-1977 and 1991-2015 (table 1). In general, highland species moved downhill while lowland species moved uphill, though the magnitude of the shift was highly variable across species. *A. gundlachi* experienced the largest shift both in median elevation (480 to 384 meters) and lower elevation limit (223 to 98 meters). The largest increase in maximum elevation occurred in *A. cristatellus*, which was collected at sites up to 700 Meters in the mid 20th century, and up to 912 Meters after 1991. Repeating these analyses on a per-locality basis found similar patterns (table 2), but only *A. gundlachi* shifted significantly (*p* = 0.0169, 95% CI -283 to -29 meters). 

In contrast to the patterns observed in individual species, the elevation distribution of all *Anolis* specimens did not shift significantly between time periods when compared on either a per-specimen (*p*=0.0522, 95% CI -8.54 to 5.49 meters) or per-locality (*p*=0.4105, 95% CI -25.64 to 12.47 meters) basis. Although the per-specimen analysis is nearly significant at the conventional alpha level of 0.05, the estimated shift in median elevation is a -5.54 meters, which is unlikely to explain the over 100 meter drops in median elevation observed in *A. evermanni*, *A. gundlachi*, and *A. krugi*.

```{r fig.width=3.25,fig.height=2.5}
locs <- ddply(anolis,.(locality,long,lat,ageClass,alt),summarize,n=length(day)) %>% subset(ageClass %in% c(2,4))
ggplot(data=locs,aes(alt))+facet_wrap(~ageClass)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),strip.text.x=element_text(size=6))+
  geom_density()+
  ggtitle("Anolis specimen elevation\n per locality")
```
```{r fig.width=3.25,fig.height=2.5}
a <- subset(anolis,ageClass %in% c(2,4))
ggplot(data=a,aes(alt))+facet_wrap(~ageClass)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),strip.text.x=element_text(size=6))+
  geom_density()+
  ggtitle("Anolis specimen elevation\n per specimen")

#wilcox.test(locs$alt[locs$ageClass==4],locs$alt[locs$ageClass==2],conf.int=T)
#wilcox.test(a$alt[a$ageClass==4],a$alt[a$ageClass==2],conf.int=T)
```

Figure X. Elevational distribution of all collecting effort, by age class. 

```{r warning=F}
#test for elevation shift on a per-specimen basis
ddply(subset(anolis,ageClass %in% c(2,4)),.(species),function (i){
  t <- wilcox.test(i$alt[i$ageClass==4],i$alt[i$ageClass==2],conf.int=T)
  sum <- c(pre.median=median(i$alt[i$ageClass==2]),
           pre.min=min(i$alt[i$ageClass==2]),
           pre.max=max(i$alt[i$ageClass==2]),
           post.median=median(i$alt[i$ageClass==4]),
           post.min=min(i$alt[i$ageClass==4]),
           post.max=max(i$alt[i$ageClass==4]),
           shift=t$estimate,
           p=round(t$p.value,8))
  names(sum) <- c("pre.median","pre.min","pre.max","post.median","post.min","post.max","shift","p")
  sum
})

#test for elevation shift on a per-locality basis (min 10 collections per locality). 
locs <- ddply(anolis,.(lat,long,ageClass,species,n),summarize,alt=mean(alt)) %>% subset(n>10) 
ddply(subset(locs,ageClass %in% c(2,4)),.(species),function(i){
  t <- wilcox.test(i$alt[i$ageClass==4],i$alt[i$ageClass==2],conf.int=T)
  sum <- c(pre.median=median(i$alt[i$ageClass==2]),
           pre.min=min(i$alt[i$ageClass==2]),
           pre.max=max(i$alt[i$ageClass==2]),
           post.median=median(i$alt[i$ageClass==4]),
           post.min=min(i$alt[i$ageClass==4]),
           post.max=max(i$alt[i$ageClass==4]),
           shift=t$estimate,
           p=round(t$p.value,8))
  names(sum) <- c("pre.median","pre.min","pre.max","post.median","post.min","post.max","shift","p")
  sum
})
```

Table 2. Results of wilcox tests on elevation distributions, by species and age class. 

```{r warning=F, fig.height=4, message=F,fig.width=6.5}
df <- subset(anolis,ageClass %in% c(2,4))
df$species <- "All species"
df <-rbind(df,subset(anolis,ageClass %in% c(2,4)))
ggplot(data=df,aes(x=factor(ageClass),y=alt))+facet_grid(.~species)+theme_bw()+xlab("")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),strip.text.x=element_text(size=6))+
  ggtitle("Median Report Elevation\n1952-1977 vs 1991-2015")+
  geom_point(size=.2,alpha=0.4,col="grey",position="jitter")+
  geom_boxplot(notch=F,outlier.colour = NA,fill=NA)
```

Figure 4. Boxplot of time-binned specimen elevations. Grey dots represent individual specimens. 

```{r warning=F, fig.height=4,fig.width=6, message=F}
ggplot(subset(locs,ageClass %in% c(2,4)),aes(x=alt,fill=factor(ageClass)))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),strip.text.x=element_text(size=6))+
  facet_wrap(~species)+
  scale_fill_discrete(name="Age Class")+
  geom_density(alpha=0.5,adjust=1)
```

Supplementary Figure X. Elevation distributions by species and age class.
\newpage

####Low-Elevation Species Composition

Focusing only on specimens collected from sites below 250M, we foundnd that the proportions of all species are independent in the midcentury and recent time periods (*p* < 0.01, Table X). In the case of *A. gundlachi*, the species represented less than 1% of the 3226 specimens collected below 250M from 1952-1977, and 8.2% from 1991-2015 (Table X).


```{r fig.width=2,fig.height=2.5}
anolis$altbin <- cut(anolis$alt,seq(0,max(anolis$alt),250),labels=c("0-250","250-500","500-750","750-1000"))
anolis.altbin <- subset(anolis,is.na(altbin)==F)
altsummary <- ddply(anolis.altbin,.(timebin,altbin,species),summarize,n=length(species))
ggplot(data=altsummary,aes(x=timebin,y=n,fill=altbin))+geom_bar(stat="identity")+ylab("Specimens")+
  ggtitle("All Anolis")+theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position="none")
```
```{r fig.height=2.5,fig.width=5}
ggplot(data=altsummary,aes(x=timebin,y=n,fill=altbin))+geom_bar(stat="identity")+ggtitle("Per Species")+
  facet_grid(.~species)+ylab("")+scale_fill_discrete(name="Altitude (M)")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),strip.text.x = element_text(size = 6))
```

Figure 9. Number of collections by species and time period. 

Chi-Squared Matrix|pre    |post
------------------|-------|-------
*A. gundlachi*    |6      |149
*Anolis sp*       |3226   |1813

```{r tidy=T}
low <- subset(anolis,alt<=250)
n.sp.pre <- nrow(subset(low,species=="Anolis gundlachi" & ageClass==2))
n.sp.post <- nrow(subset(low,species=="Anolis gundlachi" & ageClass==4))
n.all.pre <- nrow(subset(low,ageClass==2))
n.all.post <- nrow(subset(low,ageClass==4))
chimatrix <- matrix(c(n.sp.pre,n.sp.post,n.all.pre,n.all.post),byrow = T, nrow = 2)
chisq.test(chimatrix)
```

```{r}
anolischi2fun <- function(i){
n.sp.pre <- nrow(subset(low,species==i & ageClass==2))
n.sp.post <- nrow(subset(low,species==i & ageClass==4))
n.all.pre <- nrow(subset(low,ageClass==2))
n.all.post <- nrow(subset(low,ageClass==4))
chimatrix <- matrix(c(n.sp.pre,n.sp.post,n.all.pre,n.all.post),byrow = T, nrow = 2)
chi <- chisq.test(chimatrix)
c(i,round(n.sp.pre/n.all.pre,3),round(n.sp.post/n.all.post,3),round(chi$p.value,5))
}
chi2table <- data.frame(foreach(i=levels(factor(low$species)),.combine=rbind) %do% anolischi2fun(i),row.names=NULL)
names(chi2table) <- c("Species","relAbund.pre","relAbund.post" ,"p")
chi2table
```

Table X. Results of chi-squared tests on per-species specimen counts relative to all *Anolis* specimens from 1952-1977 and 1991-2015. 
\newpage

####Landcover

```{r warning=F,message=F}
setwd("~/Dropbox/anolis/")
library(raster)
#1. load landcover raster (Helmer et al. 2008, JGRBiogeoscience), reclassify to combine soil/forest types as needed
roads <- shapefile("./data/roads/tl_2013_72_prisecroads/tl_2013_72_prisecroads.shp")
lc <- raster("./data/landcover/prforestage_foresttype/IITF_JGR113_puertorico_fortype_age.img")
m.all <- read.delim("./data/landcover/prforestage_foresttype/combineSoils_reclassMatrix.txt",header=F)
m.forests <- read.delim("./data/landcover/prforestage_foresttype/PRlandcover_combSoils_allForest.txt",header=F)
m.humidForests <- read.delim("./data/landcover/prforestage_foresttype/PRlandcover_combSoils_humidForest.txt",header=F)
m.dryForests <- read.delim("./data/landcover/prforestage_foresttype/PRlandcover_combSoils_dryForest.txt",header=F)
lc.all <- reclassify(lc,m.all)
lc.forests <- reclassify(lc,m.forests)
lc.humidForests <- reclassify(lc,m.humidForests)
lc.dryForests <- reclassify(lc,m.dryForests)
coasts <- shapefile("~/Documents/worldclim/ne_10m_coastline/ne_10m_coastline.shp") %>% crop(ext.pr) %>% spTransform(.,crs(lc))
#writeRaster(lc,"./data/landcover/prforestage_foresttype/PRlandcover_combSoils.tif")
#lc.all classfication key:
# 0	Bare
# 1	High Density Urban 1991
# 2	New Urban and Bare 1991-2000
# 3	Low Density Development 2000
# 4	Pasture and Agriculture
# 5	Forest Age 1-9yr
# 6	Forest Age 10-22yr
# 7	Forest Age 23-49yr
# 8	Forest Age 50-64yr
# 9 Emergent Wetland
# 10  Coastal Sand/Rock
# 11  Water

#2. reclassify by age to produce rasters for <1958, 1959-1985,1986-1998,1999-2007 ("age1"-"age4")
age1 <- reclassify(lc.all,as.matrix(data.frame(from=c(0:11),to=c(0,0,0,0,0,0,0,0,1,0,0,0))))
age2 <- reclassify(lc.all,as.matrix(data.frame(from=c(0:11),to=c(0,0,0,0,0,0,0,1,1,0,0,0)))) 
age3 <- reclassify(lc.all,as.matrix(data.frame(from=c(0:11),to=c(0,0,0,0,0,0,1,1,1,0,0,0))))
age4 <- reclassify(lc.all,as.matrix(data.frame(from=c(0:11),to=c(0,0,0,0,0,1,1,1,1,0,0,0))))

par(mfrow=c(2,2),oma = c(1,1,0,0) + 0.1,mar = c(2,2,2,2) + 0.1)
plot(age1,main="1935-1951",axes=F,legend=F)
plot(coasts,lwd=.5,add=T)
plot(age2==1,main="1952-1977",axes=F,legend=F)
plot(coasts,lwd=.5,add=T)
plot(age3==1,main="1978-1990",axes=F,legend=F)
plot(coasts,lwd=.5,add=T)
plot(age4==1,main="1991-2000",axes=F,legend=F)
plot(coasts,lwd=.5,add=T)
par(mfrow=c(1,1))
```

Figure 10. Forest cover in Puerto Rico (1935-2000) based on historic aerial and satellite imagery (Helmer et al 2008).

```{r}
coords <- subset(anolis,!is.na(lat))
#3. get % reports in forested habitats for each sp. 
gundlachi.age2 <- subset(coords,ageClass %in% c(2) & species=="Anolis gundlachi")
pts2 <- SpatialPoints(data.frame(gundlachi.age2$long,gundlachi.age2$lat),proj4string = crs(proj4.wgs)) %>% spTransform(.,crs(proj4.utm))
gundlachi.age4 <- subset(coords,ageClass %in% c(4) & species=="Anolis gundlachi")
pts4 <- SpatialPoints(data.frame(gundlachi.age4$long,gundlachi.age4$lat),proj4string = crs(proj4.wgs)) %>% spTransform(.,crs(proj4.utm))
bgpts2 <- SpatialPoints(data.frame(coords$long[coords$ageClass %in% c(2)],coords$lat[coords$ageClass %in% c(2)]),proj4string = crs(proj4.wgs)) %>% spTransform(.,crs(proj4.utm))
bgpts4 <- SpatialPoints(data.frame(coords$long[coords$ageClass %in% c(4)],coords$lat[coords$ageClass %in% c(4)]),proj4string = crs(proj4.wgs)) %>% spTransform(.,crs(proj4.utm))
par(mfrow=c(2,1),oma = c(1,1,0,0) + 0.1,mar = c(2,2,2,2) + 0.1)
plot(age2,main="A. gundlachi, 1960-1977",axes=F,legend=F)+plot(coasts,add=T)+points(bgpts2,cex=.7)+points(pts2,cex=.2,col="red")
plot(age4,main="A. gundlachi, 1990-2015",axes=F,legend=F)+plot(coasts,add=T)+points(bgpts4,cex=.7)+points(pts4,cex=.2,col="red")
par(mfrow=c(1,1))
```

Figure 11. *A. gundlachi* collections (red circles) and forest cover, by time period. Open black circles are *Anolis* collecting localities in the respective time period. 

```{r warning=F,message=F}
alt.250 <- alt1s<250 & alt1s>0 
alt.250 <- reclassify(alt.250,rcl=cbind(0,NA))
alt.250 <- projectRaster(alt.250,age2)
alt.250 <- alt.250 > 0
age1.250 <- mask(age1,alt.250)
age2.250 <- mask(age2,alt.250)
age3.250 <- mask(age3,alt.250)
age4.250 <- mask(age4,alt.250)
pc.forested <- c()
for(i in list(age1.250,age2.250,age3.250,age4.250)){
  tmp <- i %>% values() %>% na.omit() %>% mean()
  pc.forested <- append(pc.forested,tmp)
}
pc.forested.all <- c()
for(i in list(age1,age2,age3,age4)){
  tmp <- i %>% values() %>% na.omit() %>% mean()
  pc.forested.all <- append(pc.forested.all,tmp)
}
#df <- data.frame(ageClass=c("1935-1951","1952-1977","1978-1990","1991-2000"),pc.forested=pc.forested)

df <- data.frame(year=c(1943,1965,1984,1995),low=pc.forested,all=pc.forested.all)
df <- melt(df,"year")
ggplot(data=df,aes(x=year,y=value,col=variable))+ylim(0,1)+geom_line()+geom_point()
```

Figure X. Change in percent forest cover, 1935-2000 (Helmer et al. 2008). 

23.5% of all recent *A. gundlachi* collecting localities are in areas that transitioned from open to forested after 1977. For gundlachi specimens below 250M, that figure is 32%. 

```{r warning=F,message=F}
gund.locs.4 <- ddply(gundlachi.age4,.(long,lat),summarize,n=length(day)) %>% 
                SpatialPoints(proj4string=crs(proj4.wgs)) %>% 
                  spTransform(proj4.utm)
gund.locs.4 <- SpatialPoints(coords=gundlachi.age4[,c("long","lat")],proj4string=crs(proj4.wgs)) %>% spTransform(proj4.utm)
               
reforested.index <- extract(age2,gund.locs.4)-extract(age4,gund.locs.4)
plot(age4)+points(gund.locs.4)+points(gund.locs.4[reforested.index==-1],col="red",cex=0.5)
length(reforested.index[reforested.index==-1])/length(reforested.index)

##### NOTE: working on getting the % low-altitude gundlachi in transition forest areas. #START HERE 28 Sept. 
gund.locs.4.low <- subset(gundlachi.age4,alt<250) %>% 
                    ddply(.(long,lat),summarize,n=length(day)) %>% 
                      SpatialPoints(proj4string=crs(proj4.wgs)) %>% 
                        spTransform(proj4.utm)
tmp <- subset(anolis,species=="Anolis gundlachi" & ageClass==4 & alt<= 250)
gund.locs.4.low <- SpatialPoints(coords=tmp[,c("long","lat")],proj4string=crs(proj4.wgs)) %>% spTransform(proj4.utm)
                 
reforested.index <- extract(age2,gund.locs.4.low)-extract(age4,gund.locs.4.low)
plot(age4)+points(gund.locs.4.low)+points(gund.locs.4.low[reforested.index==-1],col="red",cex=0.5)
length(reforested.index[reforested.index==-1])/length(reforested.index)

```

Figure X. A. gundlachi collections after 1991 (black circles). Red circles indicate localities that transitioned from open to forested habitats.

\newpage

#Resurveyed Localities
We identified two clusters of localities representing areas with over 100 specimens in each study period on mainland Puerto Rico: near the interection of PR-186 and PR-963 at an elevation range of 50-244M (i.e. "Ranger House"), and surrounding the El Verde Research Station on PR-186 roughly 3 km south of Ranger House at an elevation range of 210-457M. The proportion of *gundlachi* specimens was constant across time periods at El Verde Station, but increased from 0 to 3.2% at Ranger House.
[working on chi-squared tests for these, but can say already that gundlachi will have no significant shift at El Verde and will be trivially significant at Ranger House bc there are 0 gundlachi in the earlier period.] 

For reference, the (1km radius) area around Ranger House experienced an increase in forest coverage of roughly 20% from 1970 to 2000 (and 55% from 1951 to 2000), while El Verde was entirely forested across the study period. 

```{r  warning=F,message=F}
#clustering and ID'ing resurveyed localities
rsv <- subset(anolis,ageClass %in% c(2,4)) 
groups <- ddply(rsv,.(localityGroup),summarize,nAgeClass=length(unique(ageClass))) %>% 
  subset(nAgeClass>1) %>%
    .$localityGroup  # find groups w multiple ageClasses
rsv <- subset(rsv,localityGroup %in% groups)
groups <- ddply(rsv,.(localityGroup,ageClass),summarize,n=length(day)) %>%   #find groups w min specimens > 100 per ageClass
  ddply(.(localityGroup),summarize,min=min(n)) %>%
  subset(min>=100) %>% 
  .$localityGroup
rsv <- subset(rsv,localityGroup %in% groups) 


#check clusters
pts1 <- SpatialPoints(subset(rsv,localityGroup==groups[1])[,c("long","lat")],
                     proj4string=crs(proj4.wgs)) %>% spTransform(proj4.utm)
pts2 <- SpatialPoints(subset(rsv,localityGroup==groups[2])[,c("long","lat")],
                     proj4string=crs(proj4.wgs)) %>% spTransform(proj4.utm)
plot(crop(age4,ext.yunque))+points(pts1,col="red")+points(pts2,col="blue")
```

Figure X. Resurveyed locality groups, clustered at 1km. Red is El Verde Station, Blue is the Ranger House/lowland PR-186 area. 

```{r warning=F,message=F}
#species composition by cluster
spCounts <- ddply(rsv,.(localityGroup,ageClass,species),summarize,n=length(day))
groupEffort <- ddply(rsv,.(localityGroup,ageClass),summarize,total=length(day))
spCounts <- join(spCounts,groupEffort,by=c("localityGroup","ageClass"))
spCounts$freq <- spCounts$n/spCounts$total
spCounts$label <- strsplit(as.character(spCounts$species)," ") %>% sapply(function(e) e[2]) #strsplit's output is the worst
spCounts$localityGroup <- as.factor(spCounts$localityGroup)

#levels(spCounts$localityGroup) <- c("El Verde","Ranger House")

#get change in frequency bw ageClasses by site/species
diff <- ddply(spCounts,.(localityGroup,label),function(e) {
  if(nrow(e) ==2){
    diff <- e$freq[2]-e$freq[1]
  }
  if(nrow(e) == 1 && e$ageClass==4){
    diff <- e$freq-0
  }
  if(nrow(e) == 1 && e$ageClass==2){
    diff <- 0-e$freq
  }
  diff
})

ggplot(data=diff,aes(x=label,y=V1,fill=label))+
  ylab("Change in Occurrence Frequency,\n 1955-1977 vs 1991-2015")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  facet_wrap(~localityGroup)+
  geom_bar(stat="identity")

#reformat change in frequency for geom_segment()
diff <- ddply(spCounts,.(localityGroup,label),function(e) {
  if(nrow(e) ==2){
    start <- e$freq[1]
    stop <- e$freq[2]
  }
  if(nrow(e) == 1 && e$ageClass==4){
    start <- 0
    stop <- e$freq
  }
  if(nrow(e) == 1 && e$ageClass==2){
    start <- e$freq
    stop <- 0
  }
  c(start,stop)
})

ggplot(data=diff,aes(x=label,xend=label,y=V1,yend=V2))+
  facet_wrap(~localityGroup)+
  ylab("Change in Occurrence Frequency,\n 1955-1977 vs 1991-2015")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  geom_segment(arrow=arrow(length=unit(0.2,"cm")))

print(spCounts)
```

Figure X. Change in species proportions at resurveyed sites, 1952-1977 vs 1991-2015. 

```{r,echo=T,warning=F}
rh <- SpatialPoints(data.frame(long=-65.8254905,lat=18.3395686),proj4string = crs(proj4.wgs)) %>% 
        spTransform(proj4.utm) %>% 
          buffer(width=1000) 
extract(age4,rh) %>% unlist() %>% mean() - extract(age2,rh) %>% unlist() %>% mean()

rh <- SpatialPoints(data.frame(long=-65.8254905,lat=18.3395686),proj4string = crs(proj4.wgs)) %>% 
        spTransform(proj4.utm) %>% 
          buffer(width=1000) 
extract(age4,rh) %>% unlist() %>% mean() - extract(age1,rh) %>% unlist() %>% mean()
```

