---
title: 'Anolis Elevation Shifts: Prelim Summary'
author: "CJ Battey"
date: "May 20, 2016"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(tidy = T)
```
Background: 
Tropical montane species are thought to be particularly vulnerable to extinction under climate change because they occupy relatively narrow thermal niches and have a finite amount of higher-elevation habitat to occupy in the future. Simple models relating species' ranges to climatic conditions assume that as air temperatures increase, montane species will move uphill to track a thermal niche. However, physiological studies have found that body temperatures in many ectotherms are regulated via fine-scale habitat selection and microclimatic variation between (for example) forested and open habitats. Changes in average temperatures over time or along an altitudinal transect may thus be secondary to local variation in microclimate caused by shifts in land-use or habitat structure when predicting shifts in species' distributions. 

Reforestation of former agricultural lands in Puerto Rico over the course of the 20th century offers an opportunity to compare the impacts of changes in air temperatures and land use in driving shifts in species' ranges. Lizards in the genus *Anolis* are a model system in the fields of ecomorphology, island biogeography, and thermal physiology, and have been studied intensively in Puerto Rico since at least the 1960's. *A. gundlachi* is a forest specialist historically considered a montane species, while *A. cristatellus* is a habitat generalist found primarily at low elevations. Anecdotal reports suggest that *gundlachi* is now thriving at low-elevation sites that were inhabited only by *cristatellus* during the early 1970's, but no published study has assessed shifts in *Anolis* elevational range quantitatively. 

Below is a first pass through some of the available data attempting to answer a couple of questions in this system:  

1. Has *A. gundlachi* expanded to lower elevations since the 1970's?  
2. Are range shifts in *A. gundlachi* (if any) better explained by changes in climate or landcover?  


#*Anolis* Occurrence Data
####Species Occurrence Points
I downloaded data from GBIF.org for all vouchered museum specimens of *Anolis* from Puerto Rico, split out non-georeferenced sites, and removed localities with greater than 1km uncertainty in coordinates. For each species, I counted the number of pre- and post-1980 specimens and kept only the species with at least 200 specimens in both time periods. I then extracted the altitude of each specimen locality from a digital elevation model (DEM) at 30-arc-second resolution (~1km) and added that to the dataset. 

The cleaned dataset includes 10,961 reports across 6 species. About half the existing reports are georeferenced - mostly from the MVZ. MCZ and KU also have large collections, especially of earlier specimens, but we'll have to georeference them semi-manually before we can use them. It looks like many of the locality-only specimens are from El Verde and Ranger House so they should be easy to georeference, but we'll have to be careful to avoid bias in throwing away occurrence points that we can't georeference closely enough. For now I'm just looking at the georeferenced localities when analyzing change in altitude. Specimens are mostly from 1960-1980 or 2000-2010, which is convenient for binning time periods. For some reason 1971 and 1972 were particularly bad years to be an anole in Puerto Rico - there are close to 4,000 specimens from those two years alone.  
\newpage
```{r message=F,fig.height=4,warning=F}
##Anolis elevation shift analysis
setwd("/R/anolis/")
library(data.table);library(raster);library(ggplot2);library(plyr);library(foreach)
ext <- extent(-67.5,-65.5,17.8,18.7)

##### loading and cleaning data #####
anolis <- read.delim("data/locs/gbif_PRanolis/occurrence.txt")
anolis <- subset(anolis,anolis$year != "" 
                 & basisOfRecord=="PRESERVED_SPECIMEN"
                 & species != "") 
anolis$timebin <- factor(anolis$year < 1980)
levels(anolis$timebin) <- c("1980-2015","1955-1980")
anolis$timebin <- factor(anolis$timebin,levels=c("1955-1980","1980-2015"))
anolis$hascoords <- factor(!is.na(anolis$decimalLatitude) & !is.na(anolis$decimalLongitude))
levels(anolis$hascoords) <- c("locality","coordinates")
nocoords <- subset(anolis,hascoords=="locality")
coords <- subset(anolis,hascoords=="coordinates")

#read in altitude layer
alt <- crop(raster("data/alt_30s_bil/alt.bil"),ext)

#convert to spatial points
anolis.pts <- SpatialPointsDataFrame(data.frame(coords$decimalLongitude,coords$decimalLatitude),
                                     data=data.frame(coords$year,coords$month,coords$timebin,coords$species))

#extract altitudes & recombine with full dataset
anolis.alt <- extract(alt,anolis.pts) 
anolis$alt[anolis$hascoords=="coordinates"] <- anolis.alt


######### descriptive stats and summary plots #########
#check number of pre/post 1980 reports by species, filter for species w n reports per era
sp.reps <- ddply(anolis,.(species,timebin),summarize,reports=length(species))
common.species <- sp.reps$species[sp.reps$reports > 200]
anolis <- subset(anolis,species %in% common.species)
anolis$species <- factor(anolis$species)

par(mfrow=c(3,2),
    oma = c(1,1,0,0) + 0.1,
    mar = c(2,2,2,2) + 0.1)
for(i in levels(factor(anolis$species))){
  plot(alt,legend=F)+points(anolis.pts[anolis.pts@data$coords.species == i,],cex=.8)+title(i,line=0.2)
}
par(mfrow=c(1,1))
```

Fig 1. Georeferenced *Anolis* specimen localities. Base layer is elevation.
  

```{r fig.height=2.5}
#distribution of collections across museums
museums <- ddply(anolis,.(institutionCode,timebin,hascoords),summarize,Reports=length(species))
ggplot(data=museums,aes(x=substr(institutionCode,1,4),y=Reports,fill=timebin))+theme(axis.text.x = element_text(angle = 90, hjust =1,vjust=.5,size=6))+facet_grid(.~hascoords,scales="free_x",space="free_x")+xlab("")+
  geom_bar(stat="identity")+ylab("Specimens")+ggtitle("Specimens  per Institution")
```

```{r fig.height=2.5}
years <- ddply(subset(anolis,year>1960),.(year,species),summarize,Reports=length(alt))
ggplot(data=years,aes(x=year,y=Reports,fill=species))+theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  geom_bar(stat="identity")+ylab("Specimens")+xlab("")
```

Fig 2-3. Specimen source and time summaries.  

#Preliminary Elevation Analysis
Looking at the raw data without correcting for differences in survey effort, five of six species show significant decreases in mean elevation between time bins. *A. krugi*, *A. pulchellus*, and  *A. gundlachi* show the largest declines. If we analyze medians instead of means (see boxplots below), *A. cristatellus* also shows no significant change. Here's a summary table of t-test results comparing pre- and post-1980 elevations for each species:


```{r }
anolis <- subset(anolis,is.na(alt)==F)
ddply(anolis,.(species),function (i){
  t <- t.test(i$alt~i$timebin)
  p <- round(t$p.value,5)
  means <- round(t$estimate,digits=2)
  diff <- round(t$conf.int,digits=2)
  sum <- c(pre=means[1],post=means[2],diff=means[1]-means[2],ci=paste(round(diff[1],2),"-",round(diff[2],2)),p=p)
  names(sum) <- c("pre","post","difference","CI","p")
  sum
})
```




```{r warning=F, fig.height=3, message=F}
ggplot(data=anolis,aes(x=timebin,y=alt))+facet_grid(.~species)+theme_bw()+xlab("")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),strip.text.x=element_text(size=6))+
  ggtitle("Median Report Elevation, Pre vs. Post 1980")+
  geom_boxplot(notch=F,outlier.colour = NA)+
  geom_point(size=.5,alpha=0.1)
```

Figure 4. Boxplot of time-binned specimen elevations. Center is median, box is 25th and 75th percentile, bars give max/min excluding outliers.  

```{r warning=F, fig.height=3, message=F}
ggplot(data=subset(anolis,year>1960),aes(x=year,y=alt))+facet_grid(.~species)+theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),strip.text.x=element_text(size=6))+
  scale_x_continuous(breaks=c(1960,2000))+
  geom_point(size=1,alpha=0.3)+
  geom_smooth(method="lm")
```
Figure 5. Same data with time as a continuous variable.
\newpage

##Survey Effort
It seems likely that variation in survey effort is driving at least some of this pattern. For example, in *A. krugi* the large number of high-elevation reports collected at single localities during 1955-1980 pulls the "pre" average up to the approximate height of El Verde despite the presence of records all the way down to sea level in both periods. It's hard to know if those reports represent higher density of *krugi* at high altitudes, or higher survey effort. If we bin altitudes across all reports and split by time period, it looks like collecting effort has been more or less constant at low altitudes but has dropped substantially at high altitudes. Here's a summary across the six commonly collected species of *Anolis*:

```{r fig.width=2,fig.height=2.5}
anolis$altbin <- cut(anolis$alt,seq(0,max(anolis$alt),250),labels=c("0-250","250-500","500-750","750-1000"))
anolis.altbin <- subset(anolis,is.na(altbin)==F)
altsummary <- ddply(anolis.altbin,.(timebin,altbin,species),summarize,n=length(species))
ggplot(data=altsummary,aes(x=timebin,y=n,fill=altbin))+geom_bar(stat="identity")+ylab("Specimens")+
  ggtitle("All Anolis")+theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position="none")
```
```{r fig.height=2.5,fig.width=5}
ggplot(data=altsummary,aes(x=timebin,y=n,fill=altbin))+geom_bar(stat="identity")+ggtitle("Per Species")+
  facet_grid(.~species)+ylab("")+scale_fill_discrete(name="Altitude (M)")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),strip.text.x = element_text(size = 6))
```

One way to control some of the bias in survey effort is to restrict our analysis to elevations below 500M where the total number of specimens in both periods is fairly constant. Then we can ask: "has the relative abundance of gundlachi increased at low elevations between survey periods?" Using a chi-squared test across all species we get:  

Chi-Squared Matrix|pre    |post
------------------|-------|-------
*A. gundlachi*    |163    |393
*Anolis sp*       |1901   |1936

```{r tidy=T}
low <- subset(anolis,alt<=500)
n.sp.pre <- nrow(subset(low,species=="Anolis gundlachi" & timebin=="1955-1980"))
n.sp.post <- nrow(subset(low,species=="Anolis gundlachi" & timebin=="1980-2015"))
n.all.pre <- nrow(subset(low,timebin=="1955-1980"))
n.all.post <- nrow(subset(low,timebin=="1980-2015"))
chimatrix <- matrix(c(n.sp.pre,n.sp.post,n.all.pre,n.all.post),byrow = T, nrow = 2)
chisq.test(chimatrix)
```
  
The fraction of *A. gundlachi* collected below 500 meters has increased from around 9% in 1955-1980 to 20% in 1980-2016 (p<<0.01). This is the simplest version of what Siever Rohwer has called an "abundance index" in a few recent publications aimed at describing shifts in bird distributions in Mexico. If I loop the above analysis to compare across species, it looks like five of six show significant differences in relative abundance below 500 meters.  

```{r}
anolischi2fun <- function(i){
n.sp.pre <- nrow(subset(low,species==i & timebin=="1955-1980"))
n.sp.post <- nrow(subset(low,species==i & timebin=="1980-2015"))
n.all.pre <- nrow(subset(low,timebin=="1955-1980"))
n.all.post <- nrow(subset(low,timebin=="1980-2015"))
chimatrix <- matrix(c(n.sp.pre,n.sp.post,n.all.pre,n.all.post),byrow = T, nrow = 2)
chi <- chisq.test(chimatrix)
c(i,round(n.sp.pre/n.all.pre,3),round(n.sp.post/n.all.post,3),round(chi$p.value,5))
}
chi2table <- data.frame(foreach(i=levels(anolis$species),.combine=rbind) %do% anolischi2fun(i),row.names=NULL)
names(chi2table) <- c("Species","relAbund.pre","relAbund.post" ,"p")
chi2table
```

#Summary/Next Steps
There are still a few problems with this analysis. One thing I've been implicitly assuming is that researchers collecting specimens are equally likely to take any species they encounter in a given area. If earlier collectors were specifically targeting *cristatellus* then they may have encountered *gundlachi* at low altitudes but not collected it, which we wouldn't pick up in this dataset. It would be better to use a unit of collecting effort (something like (people x hours)/km^2) instead of the number of all *Anolis* specimens, but that information isn't easily available. If collectors took detailed enough field notes and those are still archived and available, we might be able to improve the corrections on a set of specimens. Doing the whole dataset is probably infeasible, but targeting single species (ie *gundlachi* and *cristatellus*) might be.  

There are also other ways to slice this data. We could try to identify specific localities that were surveyed in both time periods and compare relative abundance (ie, at Ranger House and El Verde). That would probably be more robust than comparing elevation bins as done above. Incorporating the non-georeferenced specimens would double the total number of records we're working with and might change the results if they have a strong geographic bias. We could also use a higher-resolution elevation grid, which should give better resolution on smaller shifts in elevation. Still, I think the results are promising for a first pass and at least show that we do have enough data to address question 1 (has *gundlachi* expanded at low altitudes?). 
\newpage

